---
title: ".mom files, \\<mom\\> tibbles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mom_mom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,    
  comment = "#>"
)
```
-----
**Work in progress**
-----

**
> a common file format is essential for you and morphometrics  
> `.mom` files and `<mom>` tibbles are the same object, presented differently
**

## Rationale for a common morphometric format

Since I started morphometrics on a daily basis, I have always been surprised that a common file format for morphometrics, to the best of my knowledge, does not exist.

Perhaps to my bigger surprise, it even does not seem to be really missing.

A common file format have direct benefits from the single user to the community scale: it is easy to share, archive, reuse, recycle, publish, etc.

In my humble opinion, morphometrics as a whole, and the questions it feeds, can hardly scale up without a common file format.

## `.mom` : yaml file and markup language

An ideal file format for morphometrics would be:

 * easy to read for humans
 * easy to parse for machines
 * easy to write and develop on

Good news, this exists : yaml (todo) (and more generally...). yaml and structure language wikipedia's pages are nice ones. 

For general i/o operations with yaml, the eponym package by xx is just perfect.

## Relations between `<mom>` and `.mom`
A `.mom` file is nothing but a yaml file with the `.mom` extension. Actually, `.mom` files are textual _and_ strutured representation of a `<mom>` tibble, no matter how small or complex it is.

Chivas looks like this in R, eg as a `<mom>` tibble:

```{r}
library(Momit)
# shorten it for the sake of readability
chivas <- chivas %>% Momocs2::coo_sample(4)
```

And like this in yaml, eg in a `.mom`:

```{r}
chivas_yaml <- chivas %>% yaml::as.yaml()
# cat help visualize what would be written in a a .mom file
chivas_yaml %>% cat()
```

Now, let's `chivas_yaml` is read back into R:
```{r}
chivas2 <- chivas_yaml %>% yaml::yaml.load() %>% tibble::as_tibble()
chivas2
```

And, finally, let's compare with the original tibble. We unsophisticate the mom for a tibble:

```{r}
chivas %>% tibble::as_tibble()
```

Not bad isn't? If you dissecate the two prints, you will notice that three things were dropped: 

1. the `mom_df` class (but we asked for it)
2. the `coo_list` nature of the `coo` column. Now it's just a `list`
3. the `names` within this column
